---
title: "IE 230: Google Maps API Data Extraction for a Factorial Experiment"
author: "Lino Licuanan"
date: "12/29/2020"
output: html_document
---

# Background

We were asked to conduct a simple full factorial experiment on a topic of interest. Our chosen experiment was duration in traffic, with factors being hour of day, day of week, week of month, and month of year.

Showing here the methodology to collect that information through `[googleway](https://github.com/SymbolixAU/googleway)`, a pretty handy R package for transacting with Google APIs for Places, Distance Matrix, Geocoding, and a couple other ones I didn't use.

# Data Extraction

Locations being used are a small sample of Ayala malls. As for the factor levels, setting it to the bare minimum because I'm too cheap for Google Cloud Platform lol.

```{r}
# Loading libraries
suppressPackageStartupMessages({
  library(tidyverse)
  library(magrittr)
  library(lubridate)
  library(googleway)
  library(googlesheets4)
  library(glue)
})

# Authenticating google services
gs4_auth(email = "linolicuanan@gmail.com")
set_key(Sys.getenv("GCP_GEOCODING_API_KEY"))

# Utility functions
parameter_set <- function(months, weeks, days, hours) {
  crossing(
    add_month_of_year = seq(0, 11, by = months),
    add_week_of_month = seq(0, 3,  by = weeks),
    add_day_of_week   = seq(0, 6,  by = days),
    add_hour_of_day   = seq(0, 23, by = hours)
  )
}

# Globals
# ayala_mall_sample_size <- 20
selected_malls <- c("Ayala Malls The 30th", 
                    "Glorietta by Ayala Malls", 
                    "Ayala Malls Feliz", 
                    "Ayala Malls TriNoma")

reference_date <- Sys.Date() + days(1)

# Getting ayala mall attributes
dim_ayala_mall.dt <- google_places(search_string = "ayala") %$% 
  results %>% as_tibble %>% 
  filter(name %in% selected_malls) %>% 
  mutate(id = row_number()) %>% 
  relocate(id)

# Getting pairwise distance matrix responses
test_parameter_set <- parameter_set(months = 12, weeks = 4, days = 7, hours = 24)

fact_ayala_matrix.dt <- test_parameter_set %>% 
  mutate(
    matrix_response = pmap_dfr(
      list(add_month_of_year, add_week_of_month, add_day_of_week, add_hour_of_day),
      function(add_month_of_year, add_week_of_month, add_day_of_week, add_hour_of_day) {
        try({
          output <-
            dim_ayala_mall.dt %$% 
            geometry %$% location %>% as_tibble %>% 
            transmute(geocode = map2(lat, lng, c)) %>% 
            pull %>% as.list %>% 
            google_distance(
              origins = ., 
              destinations = .,
              departure_time = as_datetime(
                reference_date + months(add_month_of_year) + weeks(add_week_of_month) +
                  days(add_day_of_week) + hours(add_hour_of_day)
              ),
              mode = "driving",
              traffic_model = "pessimistic"
            ) %$%
            rows %$% elements %>% 
            enframe %>% unnest(value) %>% 
            transmute(
              origin_id  = name,
              distance_m = distance$value,
              duration_s = duration$value,
              duration_in_traffic_s = duration_in_traffic$value
            ) %>% 
            group_by(origin_id) %>% 
            mutate(
              destination_id = row_number(), 
              departure_time = reference_date + months(add_month_of_year) + 
                weeks(add_week_of_month) + days(add_day_of_week) + hours(add_hour_of_day)
            ) %>% 
            ungroup() %>% 
            mutate(
              month_of_year = month(departure_time),
              week_of_month = ceiling(day(departure_time)/7),
              day_of_week   = wday(departure_time),
              hour_of_day   = hour(departure_time),
              updated_at    = Sys.time()
            ) %>% 
            relocate(origin_id, destination_id, departure_time, 
                     month_of_year, week_of_month, day_of_week, hour_of_day) %>% 
            nest
        })
      }
    )
  ) %>% 
  unnest(matrix_response) %>% unnest %>% 
  select(-starts_with("add"))

dim_ayala_mall.dt %>% 
  transmute(
    id, name, rating, business_status, formatted_address, 
    lat = geometry$location$lat, lng = geometry$location$lng
  ) %>% 
  sheet_write(
    ss = "1XpwhUN5iWTHBRPdqVCNutRtCXbQnolD7_33GCIPJGi0", 
    sheet = "dim_ayala_mall"
  )

fact_ayala_matrix.dt %>% 
  group_by(origin_id, destination_id, departure_time) %>% 
  slice(1) %>% 
  ungroup() %>% 
  filter(origin_id != destination_id) %>% 
  sheet_write(
    ss = "1XpwhUN5iWTHBRPdqVCNutRtCXbQnolD7_33GCIPJGi0", 
    # sheet = "fact_ayala_matrix"
    sheet = "dev.fact_ayala_matrix"
  )
```

# Simple Exploratory Analysis

```{r, fig.align='center'}
fact_ayala_matrix.dt %>% 
  group_by(origin_id, destination_id, departure_time) %>% 
  slice(1) %>% 
  ungroup() %>% 
  filter(origin_id != destination_id) %>% 
  ggplot(aes(x = distance_m, y = duration_in_traffic_s, colour = hour_of_day)) +
  geom_point()
```

